Performance analysis of high-resolution \\ ice sheet simulations

### 2022-09-25 paper/reviews/response.tex

\documentclass[letterpaper,final,12pt,reqno]{amsart}

\usepackage[total={6.3in,9.2in},top=1.1in,left=1.1in]{geometry}

\usepackage{times,bm,bbm,empheq,fancyvrb,graphicx,amsthm,amssymb}
\usepackage[dvipsnames]{xcolor}
\usepackage{longtable}
\usepackage{booktabs}

\usepackage{tabto}
\TabPositions{1.5cm}

\usepackage{float}

% hyperref should be the last package we load
\usepackage[pdftex,
colorlinks=true,
plainpages=false, % only if colorlinks=true
linkcolor=blue,   % ...
citecolor=Red,    % ...
urlcolor=black    % ...
]{hyperref}

\renewcommand{\baselinestretch}{1.05}

\allowdisplaybreaks[1]  % allow display breaks in align environments, if they avoid major underfull

\newcommand{\eps}{\epsilon}

\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}

\newcommand{\grad}{\nabla}
\newcommand{\Div}{\nabla\cdot}
\newcommand{\trace}{\operatorname{tr}}

\newcommand{\hbn}{\hat{\mathbf{n}}}

\newcommand{\bb}{\mathbf{b}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bbf}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bz}{\mathbf{z}}

\newcommand{\bF}{\mathbf{F}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bX}{\mathbf{X}}

\newcommand{\bxi}{\bm{\xi}}
\newcommand{\bzero}{\bm{0}}

\newcommand{\cK}{\mathcal{K}}
\newcommand{\cV}{\mathcal{V}}

\newcommand{\rhoi}{\rho_{\text{i}}}

\newcommand{\ip}[2]{\left<#1,#2\right>}

\newcommand{\maxR}{R^{\bm{\oplus}}}
\newcommand{\minR}{R^{\bm{\ominus}}}
\newcommand{\iR}{R^{\bullet}}

\newcommand{\nn}{{\text{n}}}
\newcommand{\pp}{{\text{p}}}
\newcommand{\qq}{{\text{q}}}
\newcommand{\rr}{{\text{r}}}

\newcommand{\supp}{\operatorname{supp}}
\newcommand{\Span}{\operatorname{span}}

\newenvironment{review}%
{\bigskip \par \begin{quote} \selectfont \sl}%
{\end{quote}}

\begin{document}
\title{Response to reviews of \emph{Performance analysis of high-resolution ice sheet simulations}}

\author{Ed Bueler}

\date{\today}

\maketitle

%\tableofcontents

\thispagestyle{empty}
%\bigskip

The thoughtful comments of the two reviewers are much appreciated.

\section{Responses to Reviewer 1}

\begin{review}
Let me congratulate you to this much needed performance analysis of ice sheet models, which is presented clearly and in detail in the manuscript at hand. I have only a few very minor comments to add.

As you include glaciers in your introduction (L\# 21) you could reference Clarke et al (Clarke, G. K., Jarosch, A. H., Anslow, F. S., Radić, V., \& Menounos, B. (2015). Projected deglaciation of western Canada in the twenty-first century. Nature Geoscience, 8(5), 372-377) around L\#30 as an example of a large, region scale glacier study using an explicit time stepping scheme on sub km grid sizes.
\end{review}

\noindent Thank you!  This is a good example to add, and I have done so.

\begin{review}
More style comments:

In lines \# 82 and on page 8 just befor equation (4), maybe use the word "equation" alongside the reference to an equation, as you do e.g. in L\#95 and other places, just for consistency.
\end{review}

\noindent Done.

\begin{review}
In lines \# 67, 150, 173, and maybe more places you use plural such as "our", "we", which strikes me as odd as you did the work on your own. But maybe that is only me as a non native English speaker.
\end{review}

\noindent I believe this is standard-enough usage, although I am happy to follow the Editor's guidance if it differs.  Speaking more precisely, I am trying to stick to ``inclusive we'' usage; see

\medskip
{\footnotesize \href{https://oxfordediting.com/to-we-or-not-to-we-the-first-person-in-academic-writing/}{\texttt{oxfordediting.com/to-we-or-not-to-we-the-first-person-in-academic-writing}}}

\medskip
\noindent Therefore I altered the text, away from ``we'' or ``our'', in lines 67, 113, 127, 173, and 207.  However, I kept ``us'', in the inclusive phrases ``let us assume'' and ``we see'', in lines 150, 186, and 217.

\begin{review}
In line \# 143, maybe add again the Robinson et al reference into the brackets to Figure 3(a) so that it is clear that you reference their figure.
\end{review}

\noindent Yes.  Done.

\begin{review}
Anyhow just a few comments from my side. Else the paper is, as said, crystal clear and well presented and very relevant for directing the model development efforts in a performance aware direction.
\end{review}

Thanks.

\section{Responses to Reviewer 2}

\end{document}

### 2022-09-26 paper/reviews/response.tex

\begin{review}
Anyhow just a few comments from my side. Else the paper is, as said, crystal clear and well presented and very relevant for directing the model development efforts in a performance aware direction.
\end{review}

Thanks.

\section{Responses to Reviewer 2}

\begin{review}
\textbf{Summary}

In this short manuscript, the author provides a prospectus on the benefits of adopting implicit methods for the time-discretization of the equations of ice sheet motion, and in particular how such methods may improve model efficiency relative to the use of explicit methods (which represent the current dominant, though not universal, paradigm). Beginning with some background on the relationship between ice sheet velocity (which is diagnostic) and geometry (which is prognostic) under different simplifications to the stress balance, the manuscript then goes through various scaling arguments that establish the essential amount of work needed to evolve an ice sheet model for a given length of time. The author re-establishes the classical result of fourth-power in spatial resolution effort for the shallow ice approximation, but presents the surprising result of (pessimistic) sixth-power growth for Stokes’ equations (which is quite oppressive), then shows that these resolution-dependent challenges can be significantly ameliorated with implicit schemes, which have the potential to realistically reduce computational cost to cubic in resolution (at least asymptotically).

Overall, the manuscript is a worthwhile contribution that will help to guide future model development and to justify why such developments are worth the effort. It is, in a sense, an opinion piece, so there is little to criticize with respect to scientific merit. I do have a limited set of comments that might help clarify points and answer questions that some readers may have.
\end{review}

\noindent This summary of my intent is quite accurate, and I appreciate the reviewer's effort to improve my manuscript.

\begin{review}
\textbf{Title} I don't find this title to be a particularly helpful characterization of the work. Maybe something more specific e.g. 'Asymptotic analysis of implicit versus explicit methods in ice sheet models' would be better.
\end{review}

\noindent Some thought on a revised title is worthwhile, so this comment is welcomed.  Unfortunately ``Asymptotic analysis of implicit versus explicit methods in ice sheet models'' would not clarify which is the asymptotic limit.   (It is not aspect ratio $H/L \to 0$, nor convergence $\Delta x \to 0$, nor approach to steady state $t\to \infty$, but rather number of degrees of freedom.)  Furthermore, my intent is to address \emph{performance}, as a broader issue which relates to multiple modeling choices, specifically mesh resolution, time-stepping type (explicit vs.~implicit), solver design (multigrid versus not), and even operational ice-climate coupling frequency, all over long-duration runs.

\medskip \noindent In other words, I don't see how to write a title of acceptable length without requiring the interested reader to (at least) dive into the abstract for clarification of some brief phrase such as ``performance analysis''.  (Of course, a shorter title will, itself, increase the number of such interested readers.)  A possible title is ``Performance analysis of numerical ice sheet simulations in the high-resolution limit'', and this captures the correct asymptotic, but I find the current title to be essentially as clear, and clearly briefer.  Does the Editor have a title suggestion?

\begin{review}
\textbf{L25} 'it is generally accepted...'  I am not sure that I agree that this is generally accepted.  There remain important works in Antarctica where models are run at 20km resolution or more and I’m not sure that they are ‘invalid’.  It is worth perhaps being more specific about the resolution requirements of specific tasks, e.g. capturing detailed perturbations to the grounding line or calving front, the influence of steep subglacial topography, etc.
\end{review}

\noindent I stand by the assertion, but this reasonable challenge has caused me to add some words to clarify.  The sentence in question now says:

\begin{quote}  In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows need yet finer resolution.\end{quote}

\noindent A few points about this sentence will, I think, be understood by any reader with minimal exposure to ice sheet modeling:
\begin{itemize}
\item This is a necessarily-imprecise description of practical resolution needs for ``resolving'' physical mechanisms understood to be important by a community of scientists.  This is not a mathematically or statistically based assertion.
\item The added phrase ''ice streams as fluid features'' refers to model results which are necessarily more than one grid cell wide.  As fluids are locally infinite-dimensional, you need a few cells, more than one or two across for ice streams, to capture a ``fluid feature''.
\item The added word ``it is \underline{now} generally accepted'' clearly refers to expectations in the 2020s.  There were indeed ``important'' 20km results before now, but I would disagree with any assertion in a present-day journal submission that, for example, flux across grounding lines in Antarctica is ``validly modeled'' at 20 km resolution; that would be a consequence of not trying very hard, or choosing a deficient model.
\item 10 km resolution is itself not enough for narrower features than ice streams.
\end{itemize}
With that background, I am happy if readers conclude, for example, that the author thinks that 20km or 40km Antarctic model results are no longer ``valid'' in a scientific sense.  That is what I think, and I was under the impression that this was a generally-accepted judgement.

\begin{review}
\textbf{L34}  While explicit time stepping is indeed the paradigm, the ‘very few exceptions’ ought to be referenced. While I am sure there are more, here are three papers that use a coupled and implicit geometry-non-SIA-velocity solution scheme: Gudmundsson (2013); Brinkerhoff et al. (2017); Shapero et al. (2021)
\end{review}

\noindent FIXME This reviewer comment drives me to make the most significant alterations to the paper.  The reviewer, and I suspect many readers, have never imagined an unconditionally stable implicit ice sheet model, so they don't recognize the only semi-implicit nature of various assertions of implicitness in the literature.  The SIA solver in Bueler (2016) is actually implicit, and demonstrably unconditionally stable on flat beds, in the sense that one can take a single time step of arbitrary length to construct the geometry of an ice sheet from knowledge only of the surface mass balance (accumulation/ablation function) and the bed.  This means a free-boundary problem must be solved during the time step to determine the map-plane region occupied by the ice.  Otherwise, in the cited models wherein an ``implicit'' claim is made, in fact a major aspect of the geometry at the current time step, specifically the ice-covered domain of the map-plane, is \emph{fixed} during the time step.  Only the vertical extent of the ice is updated implicitly.

\begin{review}
\end{review}

\noindent x
\begin{review}
\end{review}

\noindent x
\begin{review}
\end{review}

\noindent x
\begin{review}
\end{review}

\noindent x
\begin{review}
\end{review}

\noindent x
\begin{review}
\end{review}

\noindent x
\end{document}

### 2022-09-26 paper/reviews/response.tex

\begin{review}
\textbf{L34}  While explicit time stepping is indeed the paradigm, the ‘very few exceptions’ ought to be referenced. While I am sure there are more, here are three papers that use a coupled and implicit geometry-non-SIA-velocity solution scheme: Gudmundsson (2013); Brinkerhoff et al. (2017); Shapero et al. (2021)
\end{review}

\noindent FIXME This reviewer comment drives me to make the most significant alterations to the paper.  The reviewer, and I suspect many readers, have never imagined an unconditionally stable implicit ice sheet model, so they don't recognize the only semi-implicit nature of various assertions of implicitness in the literature.  The SIA solver in Bueler (2016) is actually implicit, and demonstrably unconditionally stable on flat beds, in the sense that one can take a single time step of arbitrary length to construct the geometry of an ice sheet from knowledge only of the surface mass balance (accumulation/ablation function) and the bed.  This means a free-boundary problem must be solved during the time step to determine the map-plane region occupied by the ice.  Otherwise, in the cited models wherein an ``implicit'' claim is made, in fact a major aspect of the geometry at the current time step, specifically the ice-covered domain of the map-plane, is \emph{fixed} during the time step.  Only the vertical extent of the ice is updated implicitly.

FIXME owe the reader a definition of ``explicit'', ``semi-implicit'', and ``implicit'' as I use them; put this at the end of the (new) ``Coupled geometry-velocity time-stepping'' section; ``explicit if all aspects of ice geometry are fixed during the velocity solve, and then that velocity is used (without further alteration) to compute an updated velocity''; ``semi-implicit if any major aspect of the ice geometry is fixed during the (partly-coupled) geometry-velocity solve'' (and the ice-covered map-plane region certainly counts as a major aspect); ``implicit if all aspects of ice geometry and velocity are solved for simultaneously'' (and this is necessarily a free-boundary problem in horizontal and vertical directions; quote ``weakly recognized'' comment from Schoof \& Hewitt

### 2022-09-28 paper/reviews/response.tex

\noindent A few points about this sentence will, I think, be understood by any reader with minimal exposure to ice sheet modeling:
\begin{itemize}
\item This is a necessarily-imprecise description of practical resolution needs for ``resolving'' physical mechanisms understood to be important by a community of scientists.  This is not a mathematically or statistically based assertion.
\item The added phrase ''ice streams as fluid features'' refers to model results which are necessarily more than one grid cell wide.  As fluids are locally infinite-dimensional, you need a few cells, more than one or two across for ice streams, to capture a ``fluid feature''.
\item The added word ``it is \underline{now} generally accepted'' clearly refers to expectations in the 2020s.  There were indeed ``important'' 20km results before now, but I would disagree with any assertion in a present-day journal submission that, for example, flux across grounding lines in Antarctica is ``validly modeled'' at 20 km resolution; that would be a consequence of not trying very hard, or choosing a deficient model.
\item 10 km resolution is itself not enough for narrower features than ice streams.
\end{itemize}
With that background, I am happy if readers conclude, for example, that the author thinks that 20km or 40km Antarctic model results are no longer ``valid'' in a scientific sense.  That is what I think, and I was under the impression that this was a generally-accepted judgement.

\begin{review}
\textbf{L34}  While explicit time stepping is indeed the paradigm, the ‘very few exceptions’ ought to be referenced.  While I am sure there are more, here are three papers that use a coupled and implicit geometry-non-SIA-velocity solution scheme: Gudmundsson (2013); Brinkerhoff et al. (2017); Shapero et al. (2021)
\end{review}

\noindent This reviewer comment drives me to make the most significant addition to the paper.  The reviewer, and I suspect many readers, have never really imagined an unconditionally-stable and fully-implicit ice sheet model, so they don't recognize the only semi-implicit nature of various assertions of implicitness in the literature.  The only such solver is the SIA implementation in Bueler (2016), though in fact the idea has existed as a theoretical matter since Calvo et al (2003).

My addition is the following definitions-of-terms inserted at the end of the second section, which has been re-titled from ``Mass continuity equation'' to ``Coupled geometry-velocity modeling'':

\begin{quote}
Precise definitions of the terms \emph{explicit} and \emph{implicit}, for coupled geometry-velocity updates, equivalently for coupled mass-momentum conservation time steps, will be needed for understanding the performance analysis which follows.  As defined earlier, an \emph{explicit} scheme for the coupled problem alternates between solving the stress balance for velocity, and then holding that velocity fixed while solving mass continuity or surface kinematical equation for geometry (thickness or surface elevation), and then accepting both fields as the scheme proceeds to the next time step.  At the other extreme a (\emph{fully}) \emph{implicit} scheme 
\end{quote}

FIXME returning to the reviewer's specific comment and citations, ...

 is actually implicit, and demonstrably unconditionally stable on flat beds, in the sense that one can take a single time step of arbitrary length to construct the geometry of an ice sheet from knowledge only of the surface mass balance (accumulation/ablation function) and the bed.  This means a free-boundary problem must be solved during the time step to determine the map-plane region occupied by the ice.  Otherwise, in the cited models wherein an ``implicit'' claim is made, in fact a major aspect of the geometry at the current time step, specifically the ice-covered domain of the map-plane, is \emph{fixed} during the time step.  Only the vertical extent of the ice is updated implicitly.

### 2022-10-06 paper/perfmod.tex
again put in review format
edits through big addition

\documentclass[twocolumn,letterpaper]{igs}
%\documentclass[twocolumn]{igs}  % A4 paper needs no option
%\documentclass[review,letterpaper]{igs}

@@

\abstract{Numerical ice sheet models compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.  At high spatial resolution, with horizontal mesh/grid resolutions of a few kilometers or smaller, these models usually require time steps shorter than climate-coupling time scales because they update ice thickness after each velocity solution.  High-resolution performance is degraded by the stability restrictions of such explicit time-stepping.  This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, attempts to clarify numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.  The performance of current-generation explicit time-stepping models is assessed, and then compared to the prospective performance of implicit schemes.  The main result, Table \ref{tab:performancemodel}, highlights the key roles played by the algorithmic scaling of stress-balance and implicit-step solvers.}

@@

Numerical ice sheet (glacier) models with evolving ice geometry are now routinely used for addressing scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.  It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows need yet finer resolution.  Whether using local mesh refinement \citep[for example]{Fischleretal2022,Hoffmanetal2018} or not, resolutions of two or one kilometers \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used for science at ice sheet scale.

Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, up to the non-shallow and non-hydrostatic Stokes approximation.  With very few exceptions, however, current time-stepping models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the just-computed velocity field.  Therefore ice thickness and surface elevation, functionally-equivalent variables for this purpose, are updated after velocity is fixed.  Note that the interaction between the ice sheet and the surrounding climate occurs during this geometry-update operation, via the mass continuity or surface kinematical equations \citep{GreveBlatter2009}.  Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes in particular.

These current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use velocity computed from the geometry at the start of the time step \citep[for example]{Chengetal2017}.}  At least for simpler, textbook partial differential equation problems, the conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.  Stability conditions of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.

However, actual \emph{implicit} time-stepping \citep{LeVeque2007} should also be considered.  Here the velocity and geometry are updated simultaneously by solving coupled mass and momentum conservation equations.  While implicit time-stepping always requires the solution of systems of equations at each step, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}, namely the 3D extent of the ice once the coupled solution has converged.  The problem is of free-boundary type, in map-plane (horizontal) directions \citep{SchoofHewitt2013}, and also in easily-resolved vertical directions.

An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016,JouvetGraeser2013}.  \cite{Bueler2016} also shows how the steady-state problem \citep{JouvetBueler2012} can often be solved, demonstrating unconditional stability.  (Observe that steady-state equations correspond to an implicit time step of infinite duration.)

@@

This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.  Our simplified performance analysis exposes the most important considerations and trade-offs.  While the author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, these beliefs should be assessed quantitatively to the extent possible.

\section{Mass continuity equation}

Reader familiarity is assumed with the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.  These (continuum) models are regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}.  Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.  However, before analyzing performance, the form of the mass continuity equation must be examined.

For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged horizontal velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, this equation says
\begin{equation}
\frac{\partial H}{\partial t} + \Divx \left(\bu H\right) = a, \label{eq:masscontinuity}
\end{equation}
where $\bx=(x,y)$ denotes horizontal coordinates.

Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.  This is because ice flows dominantly downhill.  Indeed, ice sheet flow has no characteristic curves, as would \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.

Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.  Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  Under a numerical thickness perturbation the velocity $\bu$ will often respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term is along $\gradx s$.  Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\grad_x s)$ which is a non-local function of geometry, and a stress balance solution is required to evaluate it.  A numerical instability occurs when the ice thickness under/over-shoots its correct value because the numerical velocity response from evaluating this non-local function is too strong for the given time step.

\newcommand{\nn}{\text{n}}
A diffusive description of the mass continuity equation is valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed.  Generically, across stress balance choices, for grounded ice sheets one observes that large values of $d$ tend to indicate locations of unstable mode growth if explicit time-steps are chosen too large.

>>>>

%\documentclass[twocolumn,letterpaper]{igs}
%\documentclass[twocolumn]{igs}  % A4 paper needs no option
\documentclass[review,letterpaper]{igs}

@@

\abstract{Numerical ice sheet models compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.  At high spatial resolution, with horizontal mesh/grid resolutions of a few kilometers or smaller, these models usually require time steps shorter than climate-coupling time scales because they update ice thickness after each velocity solution.  High-resolution performance is degraded by the stability restrictions of such explicit time-stepping.  This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, clarifies the scaling of numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.  The performance of current-generation explicit time-stepping models is assessed, and then compared to the prospective performance of implicit schemes.  The main result, Table \ref{tab:performancemodel}, highlights the key roles played by the algorithmic scaling of stress-balance and coupled, implicit-step solvers.}

@@

Numerical ice sheet (glacier) models with evolving ice geometry are now routinely used for addressing scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenlandic \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.  In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows need yet finer resolution.  Whether using local mesh refinement \citep[for example]{Fischleretal2022,Hoffmanetal2018} or not, resolutions of two or one kilometers \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019,Clarkeetal2015} are increasingly used for science at ice sheet scale.

Current-generation ice sheet models apply a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, up to the non-shallow and non-hydrostatic Stokes approximation.  With very few exceptions, however, current time-stepping models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the just-computed velocity field.  Therefore ice thickness and surface elevation, equivalent geometry variables for most modeling purposes, are updated after velocity is fixed.  The interaction between the ice sheet and the surrounding climate occurs during this geometry-update operation, via the mass continuity or surface kinematical equations \citep{GreveBlatter2009}.  Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes, in particular.

In other words, these current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system describing the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use velocity computed from the geometry at the start of the time step \citep[for example]{Chengetal2017}.}  However, at least for simpler, textbook partial differential equation problems, the limited and conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.  Stability conditions of explicit SIA models appeared some time ago \citep[e.g.][]{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.

However, actual \emph{implicit} time-stepping \citep{LeVeque2007} should also be considered.  Here the velocity and geometry are updated simultaneously by solving coupled mass and momentum conservation equations.  While such implicit time-stepping requires the solution of systems of equations at each step, for glaciers and ice sheets an implicit step must also \emph{simultaneously} compute the velocity field and the \emph{domain on which the velocity is defined}, namely the 3D extent of the ice once the coupled solution has converged.  The problem is of free-boundary type, especially in map-plane (horizontal) directions \citep{SchoofHewitt2013}, but also in the (easily-resolved) vertical direction.

An implicit strategy has been demonstrated at high (900m) resolution for the Greenlandic ice sheet in the simplest frozen-base, isothermal SIA case \citep{Bueler2016,JouvetGraeser2013}.  That work also shows how the steady-state SIA problem \citep[c.f.][]{JouvetBueler2012} can be solved, demonstrating unconditional stability.  (Observe that steady-state equations correspond to an implicit time step of infinite duration.)

@@

This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.  The simplified performance analysis here exposes the most important considerations and trade-offs.  While this author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, such beliefs should be assessed quantitatively to the extent possible.

\section{Coupled geometry-velocity modeling}

Reader familiarity is assumed with the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.  These (continuum) models are regarded here as end members of current-usage stress-balance approximations.  Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.  However, before analyzing the performance consequences of numerical modeling choices, the form of the mass continuity equation will be examined, and then the terms ``explicit'' and ``implicit'' will be carefully defined.

For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged horizontal velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, the mass continuity equation says
\begin{equation}
\frac{\partial H}{\partial t} + \Divx \left(\bu H\right) = a, \label{eq:masscontinuity}
\end{equation}
where $\bx=(x,y)$ denotes the horizontal coordinates.  Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.  This is because ice flows dominantly downhill.  Indeed, ice sheet flow has no characteristic curves, as would Equation \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.

Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.  Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  A numerical thickness perturbation will often cause the velocity $\bu$ to respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term is along the surface gradient $\gradx s$.  One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\grad_x s)$, a non-local function of geometry; a stress balance solution is required to evaluate this function.  A numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step.

\newcommand{\nn}{\text{n}}
A diffusive description of the mass continuity equation is valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.  Generically, across stress balance choices, for grounded ice sheets one observes that large values of $d$ tend to indicate locations of unstable mode growth if explicit time-steps are chosen too large.

FIXME big addition
